\documentclass[12pt,a4paper,openright,twoside]{book}
\usepackage[utf8]{inputenc}
\usepackage{phd-thesis}



\mainlinespacing{1.241} % line spacing in mainmatter, comment to default (1)

\begin{document}
	
\frontmatter
\input{front}

\begin{abstract}	
%Max 2000 characters, strict.

\section*{Italiano}

La digitalizzazione della sanità è vista come una evoluzione necessaria per migliorare efficienza e qualità dei sistemi sanitari, aumentando la tracciabilità e supportando il processo decisionale.
%
% L'applicazione delle tecnologie \acl{IoT} alla sanità porta all'emergere di nuovi paradigmi per il monitoraggio e la gestione in tempo reale delle condizioni dei pazienti.
%
In questo contesto, il concetto di \ac{DT} si è affermato come un'astrazione per modellare e rappresentare asset fisici, costruendone una replica software continuamente aggiornata tramite dati raccolti dal mondo reale.

Tuttavia, domini complessi come quello sanitario sono caratterizzati da asset eterogenei le cui relazioni sono cruciali per fornire una rappresentazione completa e accurata del contesto fisico.
%
Questa tesi esplora le sfide ingegneristiche legate allo sviluppo di \emph{Ecosistemi di Digital Twins (DTE)} che integrano \ac{DT} eterogenei connessi tra loro da relazioni dinamiche.
Il concetto di \acs{DTE} è proposto come layer di astrazione a supporto dello sviluppo di applicazioni intelligenti in grado di sfruttarne le capacità per acquisire una visione globale del mondo fisico e automatizzare processi.
%
Per affrontare tali sfide, il lavoro introduce soluzioni architetturali per lo sviluppo di \ac{DT} focalizzate su modularità e interoperabilità.
%
L’approccio è completato dall'introduzione di descrizioni (semantiche) per \ac{DT}, volte a supportare sia la gestione operativa che l'integrazione nei \acs{DTE}.
%
Ispirandoci alla natura basata su hypermedia del Web e al \acl{WoT}, è stato sviluppato un prototipo di \emph{\ac{HWoDT}} per l'implementazione di \acs{DTE}.
%
Infine, viene esplorato l'uso dei Sistemi Multi-Agente (MAS) come mezzo per realizzare applicazioni intelligenti e affidabili nei \acs{DTE}, chiarendo la relazione con i \ac{DT} e contribuendo all'ingegneria dei \acs{MAS} basati sul modello \acl{BDI}.
%
L'approccio proposto è validato attraverso l'analisi e lo sviluppo di scenari applicativi nel dominio sanitario.



\paragraph{Parole Chiave:} 
Digital Twin, Internet of Things, Web of Things, Cyber-Physical Systems, Artificial Intelligence, Multi-Agent Systems, Software Engineering

\glsresetall

\newpage

\section*{English}

The digitalization of healthcare is envisioned as a necessary development to improve the efficiency and quality of healthcare systems worldwide by improving traceability and supporting decision-making.
%
% The application of \acl{IoT} technologies to healthcare further leads to the emergence of new paradigms for live monitoring and management of patients' conditions.
%
In this context, the concept of \ac{DT} has emerged as a powerful abstraction to model and represent physical assets by creating a software replica which is continuously updated with data from the physical world.

Complex domains such as healthcare, however, are characterized by a variety of heterogeneous assets whose interrelations are crucial to provide a complete and accurate representation of the physical context.
%
This thesis explores the engineering challenges in the development of \emph{\acp{DTE}} integrating heterogeneous \acp{DT} connected by dynamic relationships.
We further propose the \ac{DTE} as a layer of abstraction to support the development of intelligent applications which can exploit its capabilities to acquire a comprehensive view of the physical world and automate processes.
%
To address these challenges, this work proposes architectural and engineering solutions for the development of \acp{DT} focusing on modularity and interoperability.
%
This approach is complemented by the introduction of (semantic) descriptions for \acp{DT} that can support both the operational management and integration challenges in \acp{DTE}.
%
Through an approach inspired by the hypermedia nature of the Web and the \acl{WoT} we develop a prototype for a \emph{\ac{HWoDT}} which can be used to implement \acp{DTE}.
%
Finally, we investigate the use of \acp{MAS} as a way to implement trustworthy intelligent applications in \ac{DTE}, defining their relationship with \acp{DT} and contributing to the engineering of \ac{MAS} based on the \acl{BDI} model. 
%
The proposed approach is validated through the analysis and development of application scenarios in the healthcare domain.


\paragraph{Keywords:} Digital Twin, Internet of Things, Web of Things, Cyber-Physical Systems, Artificial Intelligence, Multi-Agent Systems, Software Engineering

\end{abstract}

\glsresetall

% \begin{dedication} % this is optional
% The most profound technologies are those that disappear.
% \\
% They weave themselves into the fabric of everyday life
% \\
% until they are indistinguishable from it.
% \\
% \vspace{1em}
% -- Mark Weiser
% \end{dedication}


\begin{acknowledgements} % this is optional
Optional. Max 1 page.
\end{acknowledgements}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\tableofcontents   
\listoffigures     % (optional) comment if empty
\lstlistoflistings % (optional) comment if empty
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\mainmatter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapterOutsidePart{Introduction}
\label{chap:introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{chapters/introduction.tex}

%****************************************************************************************
%****************************************************************************************
\part{Background and Context}
%****************************************************************************************
%****************************************************************************************

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Healthcare Digitalization}
\label{chap:back:health4.0}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{chapters/background/health4.0}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{\aclp{DT}}
\label{chap:back:DT}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{chapters/background/dt}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{\aclp{MAS}}
\label{chap:back:MAS}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{chapters/background/mas}

%****************************************************************************************
%****************************************************************************************
\part{\aclp{DTE}}
%****************************************************************************************
%****************************************************************************************

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Engineering \aclp{DT}}
\label{chap:dte:egineering-dt}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{chapters/dte/engineering-dt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{\aclp{DTE}}
\label{chap:dte:dte}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{chapters/dte/dte}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Operational Management of \aclp{DTE}}
\label{chap:dte:dtc}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{chapters/dte/dtc}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Interoperable \aclp{DTE}}
\label{chap:dte:hwodt}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{chapters/dte/hwodt}

%****************************************************************************************
%****************************************************************************************
\part{Intelligent Applications with \acl{MAS}}
%****************************************************************************************
%****************************************************************************************

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Intelligent Applications in Healthcare}
\label{chap:mas:requirements}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{chapters/mas/requirements}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{\aclp{MAS} and \aclp{DT}}
\label{chap:mas:mas-dt}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{chapters/mas/mas-dt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Engineering \acs{BDI} \aclp{MAS}}
\label{chap:mas:engineering}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{chapters/mas/engineering}


%****************************************************************************************
%****************************************************************************************
\part{Applications Scenarios and Validation}
%****************************************************************************************
%****************************************************************************************

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Pharmaceutical Supply Chain}
\label{chap:val:irst}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{chapters/validation/irst}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Trauma Management}
\label{chap:val:trauma}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{chapters/validation/trauma}

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \chapter{Personnel Shifts Management}
% \label{chap:val:ER}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{chapters/validation/ER}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{\acl{ORM}}
\label{chap:val:orm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{chapters/validation/orm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapterOutsidePart{Conclusion}
\label{chap:conclusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{chapters/conclusion}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% BIBLIOGRAPHY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\backmatter

\bibliographystyle{alpha}
\bibliography{phd-thesis, bib/iot-dt-mas, bib/hwodt-sosym, bib/edtconf-semantic-rep}

\end{document}